A Formal Verification Framework for Smart Contracts Based on Symbolic Execution

### 2024-01-08 papers/static-checker/ecoop2024/ecoop2024.tex
minor
section 2

\section{Michelson}
\label{sec:background}

Michelson is a stack-based, statically typed functional programming language.
A Michelson program is given by the type $S$ of its storage, the type $P$ of its single
parameter, and a typed list of instructions for the underlying stack machine.
A program starts on a stack with one element which contains a pair of
the current storage value (taken from the blockchain) and the parameter value, as specified by the
contract invocation.
When it finishes, the stack must contain a single element with a pair
of the final storage value (to be saved on the blockchain) and a list
of operations.
These operations are syntactic commands like token transfers or
contract invocations, among other possibilities.
The program thus computes a function $P\times S \to O^* \times S$
where $O$ is the type of operations.

Besides the standard datatypes (e.g., numbers, lists, pairs, sums, functions)
with the expected operations,
Michelson supports a number of types and operations specific to blockchain applications (e.g.,
\texttt{mutez} for tokens, \texttt{address} for blockchain
addresses). The type system in simple in the sense that it does not
support quantification over types.

Listing~\ref {lst:simple-program-add} contains a simple example of a Michelson program.
\begin{lstlisting}[float=tp,captionpos=b,caption={A simple Michelson program},label={lst:simple-program-add},language=michelson]
parameter int;
storage   int;
code  { 
(*                 pair int int :: [] *)
  DUP; UNPAIR;
(*   int :: int :: pair int int :: [] *)
  COMPARE; GE; 
(*         bool :: pair int int :: [] *)
  IF {UNPAIR; ADD}{PUSH string 'Unexpected Pair'; FAILWITH}
      }
\end{lstlisting}
\atha{Explain the example program}
\atha{Explain some important instructions for reference later on:
  AMOUNT, BALANCE, FAILWITH, \dots}
\atha{Introduce entrypoints}

\atha{Add citations/references to the high-level languages.}
While contracts may be implemented directly in Michelson, it can also
be used as a compilation target. Indeed, several high-level languages
such as LIGO, SmartPy, and Liquidity have been created. These
languages provide a more readable syntax, resembling popular
programming languages, allowing developers to write Tezos smart
contracts more comfortably. Our tools are applicable to all of them
because they compile to Michelson.

>>>>

\section{Michelson}
\label{sec:background}
\lstset{language=michelson}

Michelson is a stack-based, statically typed functional programming language.
A Michelson program is given by the type $S$ of its storage, the type $P$ of its single
parameter, and a typed list of instructions for the underlying stack machine.
A program starts on a stack with one element which contains a pair of
the current storage value (taken from the blockchain) and the
parameter value as specified by the
contract invocation.
When the program finishes, the stack must contain a single element with a pair
of the final storage value (to be saved on the blockchain) and a list
of operations.
These operations are syntactic commands like token transfers or
contract invocations, among other possibilities.
The program thus computes a function $P\times S \to O^* \times S$
where $O$ is the type of operations.

Besides the standard datatypes (e.g., numbers, lists, pairs, sums, functions)
with the expected operations,
Michelson supports a number of types and operations specific to blockchain applications (e.g.,
\texttt{mutez} for tokens, \texttt{address} for blockchain
addresses). The type system is simple in the sense that it does not
support quantification over types.

Listing~\ref {lst:simple-program-add} contains a short example of a Michelson program.
\begin{lstlisting}[float=tp,captionpos=b,caption={A Michelson program},label={lst:simple-program-add}]
parameter int;
storage   int;
code { 
(*               pair int int :: [] *)
   DUP; UNPAIR;
(* int :: int :: pair int int :: [] *)
   COMPARE; GE; 
(*       bool :: pair int int :: [] *)
   IF {UNPAIR; ADD} {PUSH string 'Unexpected Pair'; FAILWITH};
(*                        int :: [] *)
   NIL operation; PAIR
}
\end{lstlisting}
This program first duplicates its input (\lstinline|DUP|), decomposes
the pair $P\times S$ into its components (\lstinline|UNPAIR|),
compares the two value on top of the stack (\lstinline|COMPARE|), and
transforms the result into a boolean (\lstinline|GE|). The
\lstinline|IF| instruction pops the boolean and either executes the
first (for true) or second (for false) list of instructions. By now
only the initial pair is on the stack. In the true branch,
\lstinline|UNPAIR| decomposes, and then \lstinline|ADD| adds the
components. In the false branch, the \lstinline|PUSH| instruction pushes
a string message on the stack. The \lstinline|FAILWITH| instruction
abort the contract execution: tokens are consumed, but the storage and
the balance of the contract remain unchanged.

We emerge from the conditional with just the result of the addition on
the stack. The \lstinline|NIL| instruction pushes an empty list and
\lstinline|PAIR| creates the requires pair of type $O^* \times S$.

Each Tezos contract maintains a balance of tokens, which can be
accessed from Michelson using the \lstinline|BALANCE|
instruction. Contract invocations (aka token transfers) also come with
tokens besides their data arguments. The \lstinline|AMOUNT|
instruction pushes the number of tokens received on the stack. They
are already included in the current balance of the contract. During the
execution of a Michelson contract, these values are fixed.

A contract may have several entrypoints. Entrypoints are encoded by
using a sum type (call ``or type'' in Michelson). Each (nested) alternative in a sum can be
given a name, which is used as the name for an entrypoint. A call to
an entrypoint automatically adds the required \lstinline|LEFT| and
\lstinline|RIGHT| constructors to the parameter.

\atha{Add citations/references to the high-level languages.}
While contracts may be implemented directly in Michelson, it can also
be used as a compilation target. Indeed, several high-level languages
such as LIGO, SmartPy, and Liquidity have been created. These
languages provide a more readable syntax, resembling popular
programming languages, allowing developers to write Tezos smart
contracts more comfortably. Our tools are applicable to all of them
because they compile to Michelson.
\lstset{language=Caml}
